---
title: "Project for Math and Stats"
author: "Kalide Endale"
date: "12/1/2021"
output: html_document
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## R Markdown

```{r}
#packages i will need for project
library(readr)
library(tidyverse)
library(dbplyr)
library(viridis)
library(usmap)
library(modelr)
library(plotly)

```


```{r}
#import dataset
dir01 <- "/Users/Kalide/Downloads/countries of the world.csv"
path01 <- file.path(dir01)
df01 <- read_csv(path01)
```


```{r}
#Exploratory data analysis
tibble(df01)
summary(df01)
```



looks like there is a decent amount of climate, literacy, industry, service, and agriculture data missing. 
```{r}
#count number of NA's in file and visualize NAs
missing.values <- df01 %>%
  gather(key = "key", value = "val") %>%
  mutate(is.missing = is.na(val)) %>%
  group_by(key, is.missing) %>%
  dplyr::summarise(num.missing = n()) %>%
  filter(is.missing == T) %>%
  select(-is.missing) %>%
  arrange(desc(num.missing))


#plot the chart of NAs
missing.values %>%
  ggplot() +
  geom_bar(aes(x=reorder(key, +num.missing), y=num.missing, fill = key), 
           stat = 'identity') + 
  coord_flip() +
  labs(x='variable', 
       y="number of missing values", 
       title='Visual of Missing Values') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme_minimal()
```

omit na's from df
```{r}
df02 <- na.omit(df01)

#revised dataframe
head(df02)
```
Correlation of variables


The factors that correlate with GDP per capita the most are net migration, infant mortality, literacy (%), phones (per 1000), climate, Birthrate, agriculture, service
```{r fig.width=10, fig.height=10}
df02.cor <- subset(df02, select = -c(Country, Region))
ggcorrplot::ggcorrplot(cor(df02.cor), tl.cex = 12) 
```

par(mfrow=c(6, 3))
colnames <- dimnames(df03)[[2]]
for (i in names(df03)) {
    d <- density(df03[,i])
    plot(d, type="n", main=colnames[i])
    polygon(d, col="red", border="gray")
}


#gdp looks right skewed
```{r}
library(ggplot2)
df02 %>%
  keep(is.numeric) %>%                     # Keep only numeric columns
  gather() %>%                             # Convert to key-value pairs
  ggplot(aes(value)) +                     # Plot the values
    facet_wrap(~ key, scales = "free") +   # In separate panels
    geom_density()                         # as density
```

The relationship of each improved after log10 gdp per capita
```{r}
ggplot(data = df02.cor, 
       aes(x = log10(`GDP ($ per capita)`),
           y = `Net migration`)) +
  geom_point() + 
  geom_smooth() + geom_smooth(method = "lm", color = "red") +
  labs(x = "log10 gdp per capita",
       y = "Net migration", 
       title = "Plot of gdp per capita and net migration")

#-----
ggplot(data = df02.cor, 
       aes(x = log10(`GDP ($ per capita)`),
           y = `Infant mortality (per 1000 births)`)) +
  geom_point() + 
  geom_smooth() + geom_smooth(method = "lm", color = "red") +
  labs(x = "log10 gdp per capita",
       y = "Infant mortality (per 1000 births)", 
       title = "Plot of gdp per capita and Infant mortality (per 1000 births)")

#-----
ggplot(data = df02.cor, 
       aes(x = log10(`GDP ($ per capita)`),
           y = `Literacy (%)`)) +
  geom_point() + 
  geom_smooth() + geom_smooth(method = "lm", color = "red") +
  labs(x = "log10 gdp per capita",
       y = "Literacy (%)", 
       title = "Plot of gdp per capita and Literacy (%)")

#----- 
ggplot(data = df02.cor, 
       aes(x = log10(`GDP ($ per capita)`),
           y = `Phones (per 1000)`)) +
  geom_point() + 
  geom_smooth() + geom_smooth(method = "lm", color = "red") +
  labs(x = "log10 gdp per capita",
       y = "Phones (per 1000)", 
       title = "Plot of gdp per capita and Phones (per 1000)")

#-----
ggplot(data = df02.cor, 
       aes(x = log10(`GDP ($ per capita)`),
           y = Climate)) +
  geom_point() + 
  geom_smooth() + geom_smooth(method = "lm", color = "red") +
  labs(x = "log10 gdp per capita",
       y = "Climate", 
       title = "Plot of gdp per capita and Climate")


#-----
ggplot(data = df02.cor, 
       aes(x = log10(`GDP ($ per capita)`),
           y = Birthrate)) +
  geom_point() + 
  geom_smooth() + geom_smooth(method = "lm", color = "red") +
  labs(x = "log10 gdp per capita",
       y = "Birthrate", 
       title = "Plot of gdp per capita and Birthrate")

#-----
ggplot(data = df02.cor, 
       aes(x = log10(`GDP ($ per capita)`),
           y = Agriculture)) +
  geom_point() + 
  geom_smooth() + geom_smooth(method = "lm", color = "red") +
  labs(x = "log10 gdp per capita",
       y = "Agriculture", 
       title = "Plot of gdp per capita and Agriculture")

#-----
ggplot(data = df02.cor, 
       aes(x = log10(`GDP ($ per capita)`),
           y = Service)) +
  geom_point() + 
  geom_smooth() + geom_smooth(method = "lm", color = "red") +
  labs(x = "log10 gdp per capita",
       y = "Service", 
       title = "Plot of gdp per capita and Service")
```


```{r}
#add logged data into dataframe
df03 <- df02.cor
df03$log10gdp_percapita <- log10(df03$`GDP ($ per capita)`)

df03
```
re-run correlation
```{r fig.width=10, fig.height=10}
df03.cor <- subset(df03, select = -c(`GDP ($ per capita)`))
ggcorrplot::ggcorrplot(cor(df03.cor), tl.cex = 12) 
```
So, the strongest correlator to log10gdp_percapita is phones (per 1000), then birth rate, then infant mortality (per 1000 births), then agriculture, then literacy (%), service


```{r}
correlation<-cor(df03.cor$log10gdp_percapita,df03.cor[, unlist(lapply(df03.cor, is.numeric))])
t(correlation)
```


Build simple model
```{r}
lm01 <- lm(log10gdp_percapita ~ `Phones (per 1000)`, data = df03)

summary(lm01)
```


```{r}
#residuals look linear, no violation of model assumptions
df03 %>%
  add_residuals(lm01, "resid") %>%
  ggplot(aes(sample=resid)) +
  geom_qq() +
  theme_minimal() +
  labs(x = "Theoretical Quantiles", 
       y = "Sample Quantilies", 
       title = "QQ Plot: Standardized residuals" )

# residuals look standardly distributed
df03 %>%
  add_residuals(lm01, "resid") %>%
  ggplot(aes(x=resid)) +
  geom_histogram(bins=20) +
  labs(x="Residuals",
       title = "Distribution of residuals") +
  theme_minimal()
```
So, the strongest correlator to log10gdp_percapita is phones (per 1000), then birth rate, then infant mortality (per 1000 births), then agriculture, then literacy (%), service

See if any other variable can be added to model
```{r}
#----- there is a clear negative relationship
df03 %>%
  add_residuals(lm01, "resid") %>%
  ggplot(aes(x=Birthrate)) +
  geom_point(aes(y=resid)) +
  labs(x="Birthrate", 
       y="Residuals",
       title = "Residuals vs. Birthrate") +
  theme_minimal()

#----- there is a clear negative relationship 
df03 %>%
  add_residuals(lm01, "resid") %>%
  ggplot(aes(x=`Infant mortality (per 1000 births)`)) +
  geom_point(aes(y=resid)) +
  labs(x="Infant mortality (per 1000 births)", 
       y="Residuals",
       title = "Residuals vs. Infant mortality (per 1000 births)") +
  theme_minimal()

#----- there is a clear negative relationship
df03 %>%
  add_residuals(lm01, "resid") %>%
  ggplot(aes(x=Agriculture)) +
  geom_point(aes(y=resid)) +
  labs(x="Agriculture", 
       y="Residuals",
       title = "Residuals vs. Agriculture") +
  theme_minimal()

#----- looks like there is some sort of pattern but its weak. most of the residuals are concentrated in one area
df03 %>%
  add_residuals(lm01, "resid") %>%
  ggplot(aes(x=`Literacy (%)`)) +
  geom_point(aes(y=resid)) +
  labs(x="Literacy (%)", 
       y="Residuals",
       title = "Residuals vs. Literacy (%)") +
  theme_minimal()

#------ doesn't look like it adds much to model
df03 %>%
  add_residuals(lm01, "resid") %>%
  ggplot(aes(x=Service)) +
  geom_point(aes(y=resid)) +
  labs(x="Service", 
       y="Residuals",
       title = "Residuals vs. Service") +
  theme_minimal()

```
Final simple model
```{r}
lm02 <- lm(log10gdp_percapita ~ `Phones (per 1000)` + `Infant mortality (per 1000 births)` , data = df03)

summary(lm02)
```

the remaining varialbes all look random.
```{r}
#----- there is a clear negative relationship
df03 %>%
  add_residuals(lm02, "resid") %>%
  ggplot(aes(x=Birthrate)) +
  geom_point(aes(y=resid)) +
  labs(x="Birthrate", 
       y="Residuals",
       title = "Residuals vs. Birthrate") +
  theme_minimal()

#----- there is a clear negative relationship
df03 %>%
  add_residuals(lm02, "resid") %>%
  ggplot(aes(x=Agriculture)) +
  geom_point(aes(y=resid)) +
  labs(x="Agriculture", 
       y="Residuals",
       title = "Residuals vs. Agriculture") +
  theme_minimal()

#----- looks like there is some sort of pattern but its weak. most of the residuals are concentrated in one area
df03 %>%
  add_residuals(lm02, "resid") %>%
  ggplot(aes(x=`Literacy (%)`)) +
  geom_point(aes(y=resid)) +
  labs(x="Literacy (%)", 
       y="Residuals",
       title = "Residuals vs. Literacy (%)") +
  theme_minimal()

#------ doesn't look like it adds much to model
df03 %>%
  add_residuals(lm02, "resid") %>%
  ggplot(aes(x=Service)) +
  geom_point(aes(y=resid)) +
  labs(x="Service", 
       y="Residuals",
       title = "Residuals vs. Service") +
  theme_minimal()

```


kitchen sink
```{r}
lm03 <- lm(log10gdp_percapita ~ ., data = df03.cor)

step(lm03)
```


```{r}
lm03 <- lm(log10gdp_percapita ~ `Net migration` + `Infant mortality (per 1000 births)` + 
    `Phones (per 1000)` + `Arable (%)` + `Crops (%)` + `Other (%)` + 
    Birthrate + Deathrate + Agriculture + Service, data = df03.cor)

summary(lm03)

extractAIC(lm03)
```

```{r}
df03 %>%
  add_residuals(lm03, "resid") %>%
  ggplot(aes(sample=resid)) +
  geom_qq() +
  theme_minimal() +
  labs(x = "Theoretical Quantiles", 
       y = "Sample Quantilies", 
       title = "QQ Plot: Standardized residuals" )
```







#------ 

should i partiion data set first?
```{r}
set.seed(2)

df04.cv <- crossv_kfold(df03.cor, k =10)
df04.cv
```
preform cross validation
```{r}
cv_qda1 <- df04.cv %>%
  mutate(
    fit = purrr::map(train,
                     ~lm(log10gdp_percapita ~ `Net migration` + `Infant mortality (per 1000 births)` + 
    `Phones (per 1000)` + `Arable (%)` + `Crops (%)` + `Other (%)` + 
    Birthrate + Deathrate + Agriculture + Service,
                         data = .)),
rmse = purrr::map2_dbl(fit, test, ~rmse(.x,.y)))

cv_qda1
```



```{r}
mean(cv_qda1$rmse)
```

build rmse function to compare rmse of models
```{r}
do_dataframe_cv1 <- function(formula) {
  df04.cv %>%
    mutate(fit = map(train, ~lm(formula, data = .)),
           rmse = map2_dbl(fit, test, ~rmse(.x,.y))) %>%
    summarize(cv_rmse = mean(rmse)) %>%
    pull(cv_rmse)
}
```


RMSE of simple model: lm(log10gdp_percapita ~ `Phones (per 1000)`, data = df03)
```{r}
do_dataframe_cv1(log10gdp_percapita ~ `Phones (per 1000)`)
```


RMSE of second model: lm(log10gdp_percapita ~ `Phones (per 1000)` + `Infant mortality (per 1000 births)`
```{r}
do_dataframe_cv1(log10gdp_percapita ~ `Phones (per 1000)` + `Infant mortality (per 1000 births)`)

```


RMSE of kitchen sink model: lm(log10gdp_percapita ~ `Net migration` + `Infant mortality (per 1000 births)` + 
    `Phones (per 1000)` + `Arable (%)` + `Crops (%)` + `Other (%)` + 
    Birthrate + Deathrate + Agriculture + Service, data = df03.cor)
```{r}
do_dataframe_cv1(log10gdp_percapita ~ `Net migration` + `Infant mortality (per 1000 births)` + 
    `Phones (per 1000)` + `Arable (%)` + `Crops (%)` + `Other (%)` + 
    Birthrate + Deathrate + Agriculture + Service)
```



greedy selection, start with creating a partion of dataset
```{r}
df05.cv <- resample_partition(df03.cor,
                                    p=c(train=0.5,
                                        valid=0.25,
                                        test=0.25))

df05.cv
```

In forward selection, we begin with an empty model (no candidate variables), and at each step, we add the variable that improves the model the most.
```{r}
step1 <- function(response, predictors, candidates, partition) {

rhs <- paste0(paste0(predictors, collapse = "+"), "+", candidates)
formulas <- lapply(paste0(response, "~", rhs), as.formula)
rmses <- sapply(formulas,
                function(fm) rmse(lm(fm, data = partition$train),
                                  data = partition$valid))
names(rmses) <- candidates
attr(rmses, "best") <- rmses[which.min(rmses)]
rmses
}
```

initalize a variable for tracking out model
```{r}
model <- NULL

```



Population + `Area (sq. mi.)` + `Pop. Density (per sq. mi.)` + 
    `Coastline (coast/area ratio)` + `Net migration` + `Infant mortality (per 1000 births)` + 
    `Literacy (%)` + `Phones (per 1000)` + `Arable (%)` + `Crops (%)` + 
    `Other (%)` + Climate + Birthrate + Deathrate + Agriculture + 
    Industry + Service

Step 1 (no variables): the best one, and the first one to be added as a predictor is 'log10(infant mortality rate (per 1,000 births))'; RMSE: 0.2654632 
```{r}
preds <- "1"
cands <- c("Population", 
           "`Area (sq. mi.)`", 
           "`Pop. Density (per sq. mi.)`", 
           " `Coastline (coast/area ratio)`", 
           "`Net migration`","`Infant mortality (per 1000 births)`", 
           "`Literacy (%)`", "`Phones (per 1000)`","`Arable (%)`",
           "`Crops (%)`", "`Other (%)`","Climate","Birthrate",
           "Deathrate","Agriculture","Industry","Service" )

s1 <- step1("log10gdp_percapita", preds, cands, df05.cv)
model <- c(model, attr(s1, "best"))

s1
```


step 2 ()
```{r}
preds <- "`Phones (per 1000)`"
cands <- c("Population", 
           "`Area (sq. mi.)`", 
           "`Pop. Density (per sq. mi.)`", 
           " `Coastline (coast/area ratio)`", 
           "`Net migration`","`Infant mortality (per 1000 births)`", 
           "`Literacy (%)`","`Arable (%)`",
           "`Crops (%)`", "`Other (%)`","Climate","Birthrate",
           "Deathrate","Agriculture","Industry","Service" )

s1 <- step1("log10gdp_percapita", preds, cands, df05.cv)
model <- c(model, attr(s1, "best"))

s1
```


```{r}
preds <- c("`Phones (per 1000)`","Birthrate")
cands <- c("Population", 
           "`Area (sq. mi.)`", 
           "`Pop. Density (per sq. mi.)`", 
           " `Coastline (coast/area ratio)`", 
           "`Net migration`","`Infant mortality (per 1000 births)`", 
           "`Literacy (%)`","`Arable (%)`",
           "`Crops (%)`", "`Other (%)`","Climate",
           "Deathrate","Agriculture","Industry","Service" )

s1 <- step1("log10gdp_percapita", preds, cands, df05.cv)
model <- c(model, attr(s1, "best"))

s1
```


```{r}
preds <- c("`Phones (per 1000)`","Birthrate","Agriculture")
cands <- c("Population", 
           "`Area (sq. mi.)`", 
           "`Pop. Density (per sq. mi.)`", 
           " `Coastline (coast/area ratio)`", 
           "`Net migration`","`Infant mortality (per 1000 births)`", 
           "`Literacy (%)`","`Arable (%)`",
           "`Crops (%)`", "`Other (%)`","Climate",
           "Deathrate","Industry","Service" )

s1 <- step1("log10gdp_percapita", preds, cands, df05.cv)
model <- c(model, attr(s1, "best"))

s1
```


```{r}
preds <- c("`Phones (per 1000)`","Birthrate","Agriculture","`Net migration`")
cands <- c("Population", 
           "`Area (sq. mi.)`", 
           "`Pop. Density (per sq. mi.)`", 
           " `Coastline (coast/area ratio)`", 
           "`Infant mortality (per 1000 births)`", 
           "`Literacy (%)`","`Arable (%)`",
           "`Crops (%)`", "`Other (%)`","Climate",
           "Deathrate","Industry","Service" )

s1 <- step1("log10gdp_percapita", preds, cands, df05.cv)
model <- c(model, attr(s1, "best"))

s1
```


```{r}
preds <- c("`Phones (per 1000)`","Birthrate","Agriculture","`Net migration`","Service")
cands <- c("Population", 
           "`Area (sq. mi.)`", 
           "`Pop. Density (per sq. mi.)`", 
           " `Coastline (coast/area ratio)`", 
           "`Infant mortality (per 1000 births)`", 
           "`Literacy (%)`","`Arable (%)`",
           "`Crops (%)`", "`Other (%)`","Climate",
           "Deathrate","Industry")

s1 <- step1("log10gdp_percapita", preds, cands, df05.cv)
model <- c(model, attr(s1, "best"))

s1
```


```{r}
preds <- c("`Phones (per 1000)`","Birthrate","Agriculture","`Net migration`","Service","Industry")
cands <- c("Population", 
           "`Area (sq. mi.)`", 
           "`Pop. Density (per sq. mi.)`", 
           " `Coastline (coast/area ratio)`", 
           "`Infant mortality (per 1000 births)`", 
           "`Literacy (%)`","`Arable (%)`",
           "`Crops (%)`", "`Other (%)`","Climate",
           "Deathrate")

s1 <- step1("log10gdp_percapita", preds, cands, df05.cv)
model <- c(model, attr(s1, "best"))

s1
```


```{r}
preds <- c("`Phones (per 1000)`","Birthrate","Agriculture","`Net migration`","Service","Industry","`Pop. Density (per sq. mi.)`")
cands <- c("Population", 
           "`Area (sq. mi.)`", 
           " `Coastline (coast/area ratio)`", 
           "`Infant mortality (per 1000 births)`", 
           "`Literacy (%)`","`Arable (%)`",
           "`Crops (%)`", "`Other (%)`","Climate",
           "Deathrate")

s1 <- step1("log10gdp_percapita", preds, cands, df05.cv)
model <- c(model, attr(s1, "best"))

s1
```


cut off model at net migration
```{r}
step_model <- tibble(index = seq_along(model),
                     variable = factor(names(model), levels = names(model)),
                     RMSE=model)


ggplot(step_model, aes(y = RMSE)) +
  geom_point(aes(x=variable)) +
  geom_line(aes(x=index)) +
  labs(title = "Stepwise Model Selection Plot", 
       x = "Predictor Variables") +
  theme_classic() +
  theme(axis.text.x = element_text(color = "black", 
                                   size = 9, angle=30,
                                   vjust=.8, hjust=0.8)) 

```


```{r}
set.seed(4)
fit_cv_t2 <- lm(log10gdp_percapita ~ `Phones (per 1000)` + 
                  Birthrate +
                  Agriculture + `Net migration`,
                data=df05.cv$train)

rmse(fit_cv_t2, df05.cv$test)
```



best predictive model for gdp per capita
```{r}
do_dataframe_cv1(log10gdp_percapita ~ `Phones (per 1000)` + 
                  Birthrate +
                  Agriculture + `Net migration`)
```

final model
```{r}
lm04 <- lm(log10gdp_percapita ~ `Phones (per 1000)` + 
                  Birthrate +
                  Agriculture + `Net migration`,
                data=df03.cor)
summary(lm04)
```


```{r}
#remaining things to do: add notations to code chunks, do simple visualization for presentation,kint 
```

